---
title: Red Wine quality exploration 
author: Sergei Silnov
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(dplyr)
library(gridExtra)
```

## Introduction
This document is explanatory data analysis of  Red Wines dataset. This dataset contains chemical/physical properties of wines, unique id-s and quality parameter marked by professionals.

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv('wineQualityReds.csv')
wt <- tbl_df(wine)

# dimensions and column names
dim(wt)
names(wt)

# Basic info (summary stats)
summary(wt)
glimpse(wine)
```

There are 1,599 records in dataset and 13 variables.

- **X** - unique id
- **pH**, **fixed.acidity** - acidity parameters. Both have median close to mean
- **volatile.acidity** - another acidity parameter
- **citric.acid** - values are between 0 and 1. It's interesting, if it possible to have values larger than 1
- **residual.sugar** - About 75% of wines have residual sugar less than 2.6.
- **chlorides** - have some outliers, because 3rd quantile is 0.09, but max is 0.611
- **free.sulfur.dioxide**, **total.sulfur.dioxide** - possibly correlated, comparable values. interesting to take a look on non-free sulfur dioxide
- **sulphates** - possibly have some outliers
- **density** - ~ 1. Have to look to histogram.
- **alcohol** - most valuable parameter for cheap wine.
- **quality** - parameter of our main interest. There are now wines with quality less than 3 and larger than 8, because this parameter is seems to be integer, only 6 options are possible. We can try to find why some wines have higher or lower quality depends on different chemical params.

# Univariate Plots Section

## Histograms
Lets look on histograms of all variables in dataset (except X)

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}

# Factorize quality, because there are only 6 different values. 
wine$factorized.quality <- factor(wine$quality)

# Quality histogram
qplot(factorized.quality, data = wine)

# Acids
grid.arrange(
  qplot(fixed.acidity, data = wine),
  qplot(volatile.acidity, data = wine),
  qplot(citric.acid, data = wine)
)

# Sugar,  sulphates, other params
grid.arrange(
  qplot(residual.sugar, data = wine),
  qplot(chlorides, data = wine),
  qplot(free.sulfur.dioxide, data = wine),
  qplot(total.sulfur.dioxide, data = wine),
  qplot(density, data = wine),
  qplot(pH, data = wine),
  qplot(sulphates, data = wine),
  qplot(alcohol, data = wine),
  ncol = 2
)
```

## Non-linear histograms


```{r echo=FALSE, message=FALSE, warning=FALSE, log10_histograms }
# histograms logarithmic scales
grid.arrange(
  qplot(alcohol, data = wine, binwidth = 0.01, xlab='alcohol, log10') +
  scale_x_log10(),
  qplot(residual.sugar, data = wine, xlab='residual.sugar, log10') +
  scale_x_log10(),
  qplot(free.sulfur.dioxide, data = wine, xlab='free.sulfur.dioxide, log10') +
  scale_x_log10(),
  qplot(total.sulfur.dioxide, data = wine, xlab='total.sulfur.dioxide, log10') +
  scale_x_log10(),
  ncol = 2
)
```

Total sulfur dioxide distriburtion looks like normal in log10 scale. I'll add log10 of total.sulfur.dioxide to dataset for future use.

```{r echo=FALSE, message=FALSE, warning=FALSE, add_log_sulfur.dioxide }
#  add new variable: log10.total.sulfur.dioxide
wine$log10.total.sulfur.dioxide <- log10( wine$total.sulfur.dioxide)
```


Lets try to add  more variables, non-free sulfur dioxide which is the difference between total and free sulfur dioxides.

## Additional variables

```{r echo=FALSE, message=FALSE, warning=FALSE, New_Vars}
#  add new variable: nonfree.sulfur.dioxide
wine$nonfree.sulfur.dioxide <-
  wine$total.sulfur.dioxide - wine$free.sulfur.dioxide

# and plot it histogram
qplot(nonfree.sulfur.dioxide, data = wine)
```

Almost the same as the total sulfur dioxide

## 95th percentile
Because there are several variables with heavy tails it's interesting to take a look on distributions without these tails.

```{r echo=FALSE, message=FALSE, warning=FALSE, Heavy_Tails}
# Define function to draw histogram of 95% percentile of parameters
# (i.e. cut right tail)
tailless <- function(parameter) {
  # Calculate quantile
  q95 <- quantile(wine[[parameter]], 0.95)
  
  # select subset of wines
  ninetyfive <- subset(wine, wine[[parameter]] <= q95 )
  
  # Plot the data
  return(qplot(ninetyfive[[parameter]],
               data = ninetyfive,
               xlab = paste(parameter, "95th quantile")))
}

# Plot histograms with removed outliers
grid.arrange(
  tailless("total.sulfur.dioxide"),
  tailless("free.sulfur.dioxide"),
  tailless("sulphates"),
  tailless("chlorides"),
  tailless("volatile.acidity"),
  tailless("residual.sugar"),
  ncol = 2
)

```

# Univariate Analysis

### What is the structure of your dataset?
There are 1599 vines with 12 features: ( "fixed.acidity", "volatile.acidity", "citric.acid", "residual.sugar", chlorides", "free.sulfur.dioxide", "total.sulfur.dioxide", "density", "pH", "sulphates", "alcohol", "quality" ) and X (id) labels

Most wines have quality 5 and 6 (Neutral?)
Only few wines have quality 8, even fewer - 3. There are no wines with quality > 8 or quality < 3. Its 

There are no wines with alc. less 8.5
Most wines have more than 9% of alcohol and less than 13%

Normal distributions:

- pH
- density 
- quality

Close to normal distribution, with some outliers in right tail:

- sulphates
- volatile acidity 
- fixed acidity 

Skewed to left distributions:

- citric acid
- residual sugar
- chlorides
- free.sulfur.dioxide
- total.sulfur.dioxide
- alcohol

### What is/are the main feature(s) of interest in your dataset?
Most interesting in this dataset is quality of wine and and basic chemical characteristics (alcohol, pH, acids, sulphates)

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
I think it will be interesting to figure out why good wine is good and why bad wine is bad. Because most of wines have average quality - 5 and 6, I'm going to look in detail to wines with low quaility - 3,4 and high - 7,8.

### Did you create any new variables from existing variables in the dataset?
I added non-free sulphur dioxide, but it seems that it is not very helpful.
Also I added total sulphur dioxide in log10 scale, because it may be interesting for future investigation.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

There we a number of parameters with heavy tails. Replotting them without 5% of tail values, allowed to understand better real distribution.

#### Long tails
I draw histograms of 95th percentile of total sulfur dioxide, sulphates, chlorides, volatile acidity and residual sugar.
Sulphates, volatile acidity, fixed acidity have distribution close to normal. Removing 5% of largest values made distributions much closer to normal.
Same procedure for chlorides, made histogram normal. It is interesting to take a look on these outlier.

#### Citric acid
It seems that citric acid have a number of zeros. Lets calculate number of zeros and percent of such records

```{r echo=FALSE, message=FALSE, warning=FALSE, Citric_Acid_Zeros}
# Filter dataset, where citric acid is 0
citric0 <- subset(wine, citric.acid == 0)

# Calculate number of records
dim(citric0)

# percent of 
dim(citric0) / dim(wine)
```

There are 132 records (8,2% of wines) that have citric acid equal to zero.
According to this [artice](https://en.wikipedia.org/wiki/Acids_in_wine):

> Citric acid is often added to wines to increase acidity, complement a specific flavor or prevent ferric hazes. It can be added to finished wines to increase acidity and give a “fresh” flavor.

So it is interesting to review dependence between citric acid and quality of wine.

#### Log10 scale

I plotted several variable in log10 scale. After that I figured out that total sulfur dioxide distriburtion looks like normal in this scale.

# Bivariate Plots Section

Let's start with correlations between our parameters.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
  # Calc correlation for all numeric com
  cor(wine[sapply(wine, is.numeric)])
  
```

Most parameters seems to be uncorrelated. There are some correlations ~ 0.67 between **fixed.acidity** and **citric.acid**, almost same correlation between **free.sulfur.dioxide** and **total.sulfur.dioxide**.
Anticorrelation ~ 0.68 is here between **pH** and **fixed.acidity**.
Highest corelation of quality with alcohol is just 0.48. So, there are no direct dependence of quality from one phys-chem parameter of wine. 

Lets take a look on mentioned dependecies

```{r echo=FALSE, message=FALSE, warning=FALSE, High_Cor}
qplot(x = free.sulfur.dioxide, y = total.sulfur.dioxide, data = wine)
qplot(x = pH, y = fixed.acidity, data = wine)
qplot(x = pH, y = log10(wine$fixed.acidity), data = wine)
qplot(x = fixed.acidity, y = citric.acid, data = wine)
```

First two graphs didn't say something new, correlation is clearly visible.
As mentioned in [Wikipedia](https://en.wikipedia.org/wiki/PH):

> pH is defined as the decimal logarithm of the reciprocal of the hydrogen ion activity

So it should have better correlation with log10 of fixed.acidity.

```{r echo=FALSE, message=FALSE, warning=FALSE, pH_Acids}

# add log acitidity to dataset and plot
wine$log10.fixed.acidity <- log10(wine$fixed.acidity)
qplot(x = pH, y = log10.fixed.acidity, data = wine)

# Calculate correlation
with(wine, cor(pH, log10.fixed.acidity))
```

Now anticorrelation even higher: 0.706


```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_Sugar}
  
qplot(x = alcohol, y = residual.sugar, data = wine)
  
```

Although there are no good correlation, if we look on outliers from main group, it's easy to see that sweet wines have less alcohol, and opposite wines which contain more alcohol, have lower sugar, because this sugar is converted to alcohol during wine preparation. One outlier there - wine which contain 14,9% of alcohol and ~ 8% of sugar. May some additional alcohol were added to this wine during production.


Because quality is the most interested feature, let's take a look on quality plots
```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Vs_Other}

# Shortcut for boxplots
boxplotquality <- function(parameter) {
  # Plot the data
  return(qplot(x = factorized.quality,
               y = wine[[parameter]],
               xlab = 'quality',
               ylab = parameter,
               data = wine,
               geom='boxplot'))
}

# Plot boxplots
grid.arrange(
  boxplotquality("alcohol"),
  boxplotquality("total.sulfur.dioxide"),
  ncol = 2
)
grid.arrange(
  boxplotquality("free.sulfur.dioxide"),
  boxplotquality("sulphates"),
  ncol = 2
)
grid.arrange(
  boxplotquality("chlorides"),
  boxplotquality("volatile.acidity"),
  ncol = 2
)
grid.arrange(
  boxplotquality("residual.sugar"),
  boxplotquality("density"),
  ncol = 2
)
grid.arrange(
  boxplotquality("log10.fixed.acidity"),
  boxplotquality("citric.acid"),
  ncol = 2
)

  
```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?
Quality has low correlation with other parameters, it has some correlation with alcohol, it's fun, but this correlation is not strong.
On boxplot quality vs citric acid is easy to find that wines, that contain more citric acid, but wines with extreme amount of this acid have low quality.
Also there are obvious dependency of quality on **suplphates**: best wines have larger median value, and anti-correlation on **volatile.acidity**, it low for best wines. 

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
I found good negative relationship between logarthithm of fixed acidity and pH. Although it's expectable from pH formula.

### What was the strongest relationship you found? 
There are some correlations ~ 0.67 between fixed.acidity and citric.acid, almost same correlation between free.sulfur.dioxide and total.sulfur.dioxide. Not very strong. 
Another moment, that best wines have slightly higher average alcohol.
I didn't mention correlations between **nonfree.sulfur.dioxide**,  **log10.total.sulfur.dioxide** and **total.sulfur.dioxide**, because first two features were built using last one.
As mentioned earlier best correlation between log10 of fixed.acidity and **pH** = *0.706*

# Multivariate Plots Section

Start with scatter plots on features that have visibe influence to quality:  **volatile.acidity**, **suplphates**, **alcohol**, **citric.acid**

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
# Shortcut for scatter plots with quality
biscatter <- function(x, y) {
  # Plot the data
  return(qplot(x = wine[[x]],
               y = wine[[y]],
               xlab = x,
               ylab = y,
               color = factorized.quality,
               data = wine) + 
    scale_colour_brewer(palette = "Spectral", direction = -1) +
    theme_dark())
}

biscatter("volatile.acidity","citric.acid")
biscatter("sulphates","citric.acid")
biscatter("sulphates","alcohol")
biscatter("volatile.acidity","alcohol")

```

Now take a look on same plots, but only for good and bad wines, where quality < 5 and > 6

```{r echo=FALSE, message=FALSE, warning=FALSE, Good_And_Bad}

# Save new dataset to variable
gbwine <- subset(wine, (quality < 5) | (quality > 6))

# Dimentions of new dataset
dim(gbwine)
```

There are 280 items in dataset without average wines

```{r echo=FALSE, message=FALSE, warning=FALSE, Good_And_Bad_Plots}

# Shortcut for scatter plots with quality without 5 and 6
biscattergb <- function(x, y) {
  # Plot the data
  return(qplot(x = gbwine[[x]],
               y = gbwine[[y]],
               xlab = x,
               ylab = y,
               color = factorized.quality,
               data = gbwine) + 
    scale_colour_brewer(palette = "Spectral", direction = -1) +
    theme_dark())
}

biscattergb("volatile.acidity","citric.acid")
biscattergb("sulphates","citric.acid")
biscattergb("sulphates","alcohol")
biscattergb("volatile.acidity","alcohol")
biscattergb("citric.acid","alcohol")

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?
When I plotted quality data for 4 previously selected features **volatile.acidity**, **suplphates**, **alcohol** and **citric.acid** on quality, I found that on multivariate graphs dependencies are clear. Because main area of my interest were dependence on quality, I filtered out wines quality 5 and 6 to clearly see when which wines are good, and which are bad. And last bunch of plots have clearly visible groups of different parameters for good and bad wines.

### Were there any interesting or surprising interactions between features?
It is interesting that amounts of different acids: citric, acetic and tartaric determine taste and quality of wine. For example best wines tends to have citric acid 0.25-0.5 g/dm^3, but volatile  acidity (acetic acid) less than 0.4 g/dm^3. And increasing of citric acid will not make wine bad, but it's easy if there is a lot of acetic acid in wine.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
qplot(factorized.quality, data = wine, fill = I('#099DD9')) +
  ggtitle('Histogram of quality of  wines') +
  xlab('Quality') + 
  ylab('Number of wines')

```

### Description One
Distribution of quality of wine. Very close to normal. Experts avoid using low < 3 and high > 8 marks. Why are there not present ratings 0-2 and 9-10? I think it's possible that expert will rank some wine with rating 1 or 9, but this ratings are medians of at least 3 evaluations made by wine experts. and chances that all experts will put same (very high or very low) score is really small. As for me this fact demonstrates, that wine quality metric highly depends on personal taste, so it will be hard to find some strong correlations or build analytical model.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
title <- paste('Sulphates and volatile acidity over quality',
               '(with alcohol) for good and bad wines')
qplot(x = gbwine$sulphates,
      y = gbwine$volatile.acidity,
      xlab = "Sulphates, g/dm^3",
      ylab = "Volatile acidity, g/dm^3",
      size = alcohol,
      color = quality,
      data = gbwine) + 
    scale_colour_gradientn(colours=rainbow(2)) +
    scale_size_continuous(name = "alcohol, %") +
    ggtitle(title) 
```

### Description Two
This plot show 4 characteristics of selected wines (where quality 4 and less or 7 and more).
It's clear that good wines usually have lower volatile acidity with higher amount of sulphates. Another noticable moment, that bad wines have smaller amount of alcohol. 

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
bplotfinal <- function(y, label) {
  # Plot the data
  return(qplot(x = factorized.quality,
               fill = factorized.quality,
               y = wine[[y]],
               xlab = 'Quality',
               ylab = label,
               data = wine,
               geom='boxplot') +
         guides(fill=FALSE)) 
}

grid.arrange(
  bplotfinal('volatile.acidity', 'Volatile Acidity (g/dm^3)'),
  bplotfinal('citric.acid', 'Citric acid (g/dm^3)'),
  bplotfinal('sulphates', 'Sulphates (g/dm^3)'),
  bplotfinal('alcohol', 'Alcohol (% by volume)'),
  ncol = 2,
  top='Effect of volatile acidity, sulphates, alcohol and citric.acid on wine quality')
```

### Description Three
All these 4 paramaters can provide some invormation about quality of wine. So, to be the best wine should:

* have not very high density with not very high ~ 0.75 g/dm^3 of sulphates.
* have volatile acidity less than 0.4 g/dm^3.
* have quite high amount of alcohol (> 11%)
* and ~ 0.75 g/dm^3 of sulphates.

Although it's not so easy to build model of bad wine, because parameters have larger diversity for bad wines.

------

# Reflection
Dataset contains information about 1599 probes of wines with some information about quality marked by experts. Data is full without visible errors and mistakes. Athough there are not too much correlations between characteristics, and most of them are not directly related to quality, it's possible to find some differencies between good and bad wines, while there are not enough information to say something important about average wine.

I think it will be hard to build some analytical model to predict quality of wine, because paramemers be not linearly separated, and there is large component of personal taste in quality rating. Anyway, it is possible to get some intuition about quality of wine based on amount of acids, sulphates and alcohol in each example. 

I was interesting to realize that fixed acidity (tartaric acid, main component of wine acidity) in log10 scale have good anti-correlation with pH, which is log10 characteristic too.

I never thought about acids as important part of wine taste. It's interesting what happen if I add some amount of lemon juice (where 6-8% of citric acid)  to red wine with low acidity. It will be my next experiment.
